Orienting the causal relationship between imprecisely measured traits using genetic instruments

### 2017-03-11 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using genetic instruments}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{05 March 2017}

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding (1--3). In response to this, genetic instrumentation has
emerged as a technique for improving the reliability of causal inference
in observational data, and with the coincident rise in genome-wide
association studies it is now a prominent tool that is applied in
several different guises (3--6). However, potential pitfalls remain and
one that is often neglected is the influence of non-differential
measurement error on the reliability of causal inference.

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or levels of imprecision (noise) of the measurement. Such
variability can arise through a whole plethora of mechanisms, which are
often unique to the study design and difficult to avoid (7,8). Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but comes with the problem of having imperfect resolution, for
instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true value due to
imperfect sensitivity (9,10). Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types (11).

@@

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that in this simple three-variable scenario the indirect
influence of the SNP on the outcome will disappear when conditioning on
the exposure. Here, the exposure `mediates' the association between the
SNP and the outcome, providing information about the causal influence of
the exposure on the outcome. This forms the basis of a number of methods
such as genetical genomics (17), the regression-based causal inference
test (CIT) (4,18), a structural equation modelling (SEM) implementation
in the NEO software (5), and various other methods including Bayesian
approaches (6). They have been employed by a number of recent
publications that make causal inferences in large scale `omics datasets
(6,19--23).

@@

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. Genome-wide association studies (GWASs) that
identify genetic associations for complex traits are, by design,
hypothesis free and agnostic of genomic function, and it often takes
years of follow up studies to understand the biological nature of a
putative GWAS hit (27). If multiple instruments are available for an
hypothesised exposure, which is increasingly typical for complex traits
that are analysed in large GWAS consortia, then techniques can be
applied to mitigate these issues (16). But these techniques cannot
always be applied in the case of determining causal directions between
'omic measures where typically only one cis-acting SNP is known. For
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
The mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples (28). This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated (26). It is important to note that if there are
any confounders influencing both the exposure and outcome, the
mediation-based approach of adjusting the outcome for the exposure to
nullify the association between the SNP and the outcome will not work.
This is because adjusting the outcome by the exposure induces a collider
effect between the SNP and outcome (29), and the in order to fully
abbrogate this association one must also adjust for all (hidden or
otherwise) confounders. MR does not suffer from this problem because it
does not test for association through adjustment. Additionally, when MR
assumptions are satisfied the method is robust to there being
measurement error in the exposure variable (30). Indeed instrumental
variable (IV) analysis was in part initially introduced as a correction
for measurement error in the exposure (31), whereas it has been noted
that both classic mediation-based analyses (13,14,32,33) and
mediation-based methods that use instrumental variables (34,35) are
prone to be unreliable in its presence.

@@

We can model a system whereby some exposure \(x\) has a causal influence
\(\beta_x\) on an outcome \(y\) such that

@@

The \(\alpha_*\) terms represent intercepts, and the \(\epsilon_*\)
denote random error. Mediation-based analyses that test whether \(x\)
causally relates to \(y\) impinge on evaluating if the influence of
\(g\) on \(y\) can be accounted for by conditioning on \(x\), such that

@@

where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). MR analysis
estimates the causal influence of \(x\) on \(y\) by using the instrument
as a proxy for \(x\), such that

@@

Measurement error of an exposure can be modeled as a transformation of
the true value that leads to the observed value, \(x_o = f(x)\). For
example, following Pierce and VanderWeele (30) we can define

@@

where \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\). The
same model of measurement error can be applied to the outcome variable
\(y\). In this study we assume there is no measurement error in the SNP
and measurement error in the exposure and the outcome are uncorrelated.

@@

The CIT method (4) is implemented in the R package \emph{R/cit} (18).
The methodology of the CIT is as follows. Assume an exposure \(x\) is
instrumented by a SNP \(g\), and the exposure \(x\) causes an outcome
\(y\), as described above. The following tests are then performed:

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. The 4th
condition is formulated as an equivalence testing problem that is
estimated using simulations, comparing the estimate against from the
data against empirically obtained estimates for simulated variables
where the independence model is true (full details are given in (4)). We
note here that this approach is liable to fail, even when there is a
true causal relationship, when confounders are present due to inducing
collider effects.

@@

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT for each of two tests -
\(p_{CIT, x \rightarrow y}\), was estimated for the direction of \(x\)
causing \(y\) (Model 1), and for the direction of \(y\) causing \(x\),
\(p_{CIT, y \rightarrow x}\) (Model 2). The results from each of these
methods can then be used in combination to infer the existance and
direction of causality. For some significance threshold \(\alpha\) there
are four possible outcomes from these two tests, and their
interpretations are as follows:

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the R package \(R/systemfit\) (38).

@@

and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
indicate the probability of obtaining the observed difference in
correlations \(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis
that both correlations are identical.

The existence of causality and its direction is inferred based on
combining information from the MR analysis and the Steiger test. The MR
analysis indicates whether there is a potential causal relationship
(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
of the causal relationship and the confidence of the direction
(\$p\_\{Steiger\}). For the purposes of compiling simulation results,
these can be combined using an arbitrary \(\alpha = 0.05\) value:

\begin{itemize}
\tightlist
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
  a causal association for the correct model is accepted
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
  a causal association for the incorrect model is accepted
\item
  Otherwise no evidence for a causal relationship
\end{itemize}

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
because \(\rho_{g,x_o}\) could be underestimated more than
\(\rho_{g,y_o}\).

As we show in Appendix 2 it is possible to infer the bounds of
measurement error on \(x_o\) or \(y_o\) given known genetic
associations. The maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\), because it is known that at least that much of the
variance has been explained in \(x_o\) by \(g\). The minimum is 0,
denoting perfectly measured trait values (the same logic applies to
\(y_o\)). It is possible to simulate what the inferred causal direction
would be for all values within these bounds.

@@

\subsection{Two sample MR}\label{two-sample-mr}

@@

All analysis was performed using the R programming language (42) and
code is made available at \href{}{github url to go here} and implemented
in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
when the true model is causal only when \(D = 1\). Therefore, if there
is any measurement error that incurs imprecision (i.e.
\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
between \(g\) and \(y_o | x\), which is in violation of the the 4th
condition of the CIT. Note that scale transformation of \(x\) or \(y\)
without any incurred imprecision is insufficient to lead to a violation
of the test statistic assumptions, and henceforth mention of measurement
error will relate to imprecision unless otherwise stated.

@@

We performed simulations to compare the performance of MR against CIT in
detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 3
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. When MR assumptions are satisfied, notably that it is
known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
the performance of MR in detecting an association is unrelated to
measurement error in the exposure. Measurement error in the outcome does
reduce power, but does not induce a substantive difference in
performance between CIT and MR.

@@

We used simulations to explore the performance of the MR Steiger
approach in comparison to CIT in terms of the rate at which a strong
causal relationship is obtained for the correct direction of causality,
and the rate at which a strong causal relationship is obtained where the
reported direction of causality is incorrect. Simulations were performed
for two models, one for a ``causal model'' where there was a causal
relationship between \(x\) and \(y\); and one for a ``non-causal model''
where \(x\) and \(y\) were not causally related, but had a confounded
association induced by the SNP \(g\) influencing a confounder variable
\(u\).

Figure 5a shows that, for the ``causal model'', when \(d < 0\) the MR
analysis is indeed liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, the CIT is in general more fallable to reporting a
robust causal association for the wrong direction of causality. When
\(d > 0\) is satisfied we observe that in most cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality. The CIT, unlike the Steiger
test for MR, is able to distinguish the ``non-causal model'' from the
``causal model'' (Methods, Figure 5b), but it is evident that
measurement error will often lead the CIT to evaluate erroneously this
confounding model as being causal.

@@

We used the Steiger test to infer the direction of causality between 458
putative associations between DNA methylation levels and gene expression
levels and found that the causal direction commonly goes in both
directions (Figure 6a), but assuming no or equal measurement error, DNA
methylation levels were the predominant causal factor
(\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
was 3.92 (95\% CI 1.08 - 37.11). We then went on to predict the causal
directions of the associations for varying levels of systematic
measurement error for the different platforms. Figure 6a shows that the
qualitative result of which of DNA methylation or gene expression is
more commonly the causal factor is very strongly determined by
measurement error.

We performed two sample MR (28) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (41) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even for relatively low level
associations it is important to use causal inference methods over
observational associations to infer causal effect sizes.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored (44). Our analysis extends this to
related methods such as CIT that are widely used in predominantly 'omic
data. These methods are indeed liable to the same issues as standard
mediation based analysis, and specifically we show that as measurement
error in the exposure variable increases, CIT is likely to have reduced
statistical power, and liable to infer the wrong direction of causality.
We also demonstrate that, though unintuitive, increasing sample size
does not resolve the issue, rather it leads to more extreme p-values for
the model that predicts the wrong direction of causality.

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using genetic instruments}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{11 March 2017}

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding (1--3). In response to this, the use of genetic
associations to instrument traits has emerged as a technique for
improving the reliability of causal inference in observational data, and
with the coincident rise in genome-wide association studies it is now a
prominent tool that is applied in several different guises (3--6).
However, potential pitfalls remain and one that is often neglected is
the influence of non-differential measurement error on the reliability
of causal inference.

Measurement error is the difference between the measured value of a
quantity and its true value. This study focuses specifically on
non-differential measurement error where all strata of a measured
variable have the same error rate, which can manifest as changes in
scale or measurement imprecision (noise). Such variability can arise
through a whole plethora of mechanisms, which are often specific to the
study design and difficult to avoid (7,8). Array technology is now
commonly used to obtain high throughput phenotyping at low cost, but
comes with the problem of having imperfect resolution, for instance
methylation levels as measured by the Illumina450k chip are prone to
have some amount of noise around the true value due to imperfect
sensitivity (9,10). Relatedly, if the measurement of biological interest
is the methylation level in a T cell, then measurement error of this
value can be introduced by using methylation levels from whole blood
samples because the measured value will be an assay of many cell types
(11).

@@

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that in this simple three-variable scenario the estimated direct
influence of the SNP on the outcome will be zero when conditioning on
the exposure. Here, the exposure completely mediates the association
between the SNP and the outcome, providing information about the causal
influence of the exposure on the outcome. This forms the basis of a
number of methods such as genetical genomics (17), the regression-based
causal inference test (CIT) (4,18), a structural equation modelling
(SEM) implementation in the NEO software (5), and various other methods
including Bayesian approaches (6). They have been employed by a number
of recent publications that make causal inferences in large scale `omics
datasets (6,19--23).

@@

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the SNP being robustly associated with each, it can be difficult to
determine which of the two variables is subject to the primary effect of
the SNP (i.e.~for which of the two variables is the SNP a valid
instrument? Figure 1). By definition, we expect that if the association
is causal then a SNP for the exposure will be associated with the
outcome, such that if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit (27). If multiple
instruments are available for an hypothesised exposure, which is
increasingly typical for complex traits that are analysed in large GWAS
consortia, then techniques can be applied to mitigate these issues (16).
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level (Figure 1)?

MR has some important advantages over the mediation-based approaches.
First, the mediation-based approaches require that the exposure, outcome
and instrumental variables are all measured in the same data, whereas
recent extensions to MR circumvent this requirement, allowing causal
inference to be drawn when exposure variables and outcome variables are
measured in different samples (28). This has the crucial advantage of
improving statistical power by allowing analysis in much larger sample
sizes, and dramatically expands the breadth of possible phenotypic
relationships that can be evaluated (26). Second, the mediation-based
approach of adjusting the outcome for the exposure to nullify the
association between the SNP and the outcome is affected by unmeasured
confounding of the exposure and outcome. This is because adjusting the
outcome by the exposure induces a collider effect between the SNP and
outcome (29), and the in order to fully abrogate this association one
must also adjust for all (hidden or otherwise) confounders. MR does not
suffer from this problem because it does not test for association
through adjustment. Third, when MR assumptions are satisfied the method
is robust to there being measurement error in the exposure variable
(30). Indeed instrumental variable (IV) analysis was in part initially
introduced as a correction for measurement error in the exposure (31),
whereas it has been noted that both classic mediation-based analyses
(13,14,32,33) and mediation-based methods that use instrumental
variables (34,35) are prone to be unreliable in its presence.

@@

We model a system whereby some exposure \(x\) has a causal influence
\(\beta_x\) on an outcome \(y\) such that

@@

The \(\alpha_*\) terms represent intercepts, and henceforth can be
ignored. The \(\epsilon_*\) terms denote random error, assumed
independently and normally distributed with mean zero. Mediation-based
analyses that test whether \(x\) causally relates to \(y\) rely on
evaluating whether the influence of \(g\) on \(y\) can be accounted for
by conditioning on \(x\), such that

@@

where \(\hat{y} = \hat{\beta}_x x\). MR analysis estimates the causal
influence of \(x\) on \(y\) by using the instrument as a proxy for
\(x\), such that

@@

Measurement error of an exposure can be modeled as a transformation of
the true value (\(x\)) that leads to the observed value, \(x_o = f(x)\).
For example, following Pierce and VanderWeele (30) we can define

@@

where \(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\). The
same model of measurement error can be applied to the outcome variable
\(y\). In this study we assume there is no measurement error in the SNP,
and that measurement error in the exposure and the outcome are
uncorrelated.

@@

First we describe how the CIT method (4) is implemented in the R package
\emph{R/cit} (18). The methodology of the CIT is as follows. Assume an
exposure \(x\) is instrumented by a SNP \(g\), and the exposure \(x\)
causes an outcome \(y\), as described above. The following tests are
then performed:

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. The 4th
condition is formulated as an equivalence testing problem that is
estimated using simulations, comparing the estimate against from the
data against empirically obtained estimates for simulated variables
where the independence model is true (full details are given in (4)). We
note here that this approach is liable to fail, even when there is a
true causal relationship, when confounders of the exposure and outcome
are present, as these will induce collider bias.

@@

Now we describe how we used the CIT method in our simulations. The
\emph{cit.cp} function was used to obtain an omnibus p-value. To infer
the direction of causality using the CIT method, an omnibus p-value
generated by CIT for each of two tests - \(p_{CIT, x \rightarrow y}\),
was estimated for the direction of \(x\) causing \(y\) (Model 1), and
for the direction of \(y\) causing \(x\), \(p_{CIT, y \rightarrow x}\)
(Model 2). The results from each of these methods can then be used in
combination to infer the existance and direction of causality. For some
significance threshold \(\alpha\) there are four possible outcomes from
these two tests, and their interpretations are as follows:

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the \(systemfit\) function in the R package
\(R/systemfit\) (38). Note that the value of \(p_{MR}\) is identical
when using the same genetic variant to instrument the influence of the
exposure \(x\) on the outcome \(y\), or erroneously, instrumenting the
outcome \(y\) on the exposure \(x\).

@@

and a p-value, \(p_{Steiger}\) is generated from the \(Z\) value to
indicate the probability of obtaining a difference between correlations
\(\rho_{gx}\) and \(\rho_{gy}\) at least as large as the one observed,
under the null hypothesis that both correlations are identical.

The existence of causality and its direction is inferred based on
combining information from the MR analysis and the Steiger test. The MR
analysis indicates whether there is a potential causal relationship
(\(p_{MR}\)), and the Steiger test indicates the direction (\(sign(Z)\))
of the causal relationship and the confidence of the direction
(\(p_{Steiger}\)). For the purposes of compiling simulation results,
these can be combined using an arbitrary \(\alpha = 0.05\) value:

\begin{itemize}
\tightlist
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
  a causal association for the correct model is accepted,
  \(x \rightarrow y\)
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
  a causal association for the incorrect model is accepted,
  \(y \rightarrow x\)
\item
  Otherwise if \(p_{Steiger} > \alpha\) or \(p_{MR} > \alpha\), neither
  model is accepted
\end{itemize}

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) (or
\(\rho_{gy}\)) are underestimated if \(x\) (or \(y\)) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
because \(\rho_{g,x_o}\) could be underestimated more than
\(\rho_{g,y_o}\).

As we show in Appendix 2 it is possible to infer the bounds of
measurement error on \(x_o\) or \(y_o\) given known genetic
associations. The maximum measurement imprecision of \(x_o\) is
\(\rho_{g,x_o}\), because it is known that at least that much of the
variance has been explained in \(x_o\) by \(g\). The minimum is 0,
denoting perfectly measured trait values (the same logic applies to
\(y_o\)). It is possible to simulate what the inferred causal direction
would be for all values within these bounds.

@@

\subsection{Applied example using two sample
MR}\label{applied-example-using-two-sample-mr}

@@

All analysis was performed using the R programming language (42) and
code is made available at
\url{https://github.com/explodecomputer/causal-directions} and
implemented in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
when the true model is causal only when \(D = 1\). Therefore, if there
is any measurement error that incurs imprecision in \(x\) (i.e.
\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
between \(g\) and \(y_o | x_o\), which is in violation of the the 4th
condition of the CIT. Note that scale transformation of \(x\) or \(y\)
without any incurred imprecision is insufficient to lead to a violation
of the test statistic assumptions, and henceforth mention of measurement
error will relate to imprecision unless otherwise stated.

@@

We also performed simulations to compare the performance of MR against
CIT in detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 3
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. When MR assumptions are satisfied, notably that it is
known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
the performance of MR in detecting an association is unrelated to
measurement error in the exposure. Measurement error in the outcome does
reduce power, but does not induce a substantive difference in
performance between CIT and MR.

@@

We used simulations to explore the performance of the MR Steiger
approach in comparison to CIT in terms of the rate at which evidence of
a causal relationship is obtained for the correct direction of
causality, and the rate at which evidence of a causal relationship is
obtained where the reported direction of causality is incorrect.
Simulations were performed for two models, one for a ``causal model''
where there was a causal relationship between \(x\) and \(y\); and one
for a ``non-causal model'' where \(x\) and \(y\) were not causally
related, but had a confounded association induced by the SNP \(g\)
influencing a confounder variable \(u\).

Figure 5a shows that, for the ``causal model'', the MR analysis is
indeed liable to infer the wrong direction of causality when \(d < 0\),
and that this erroneous result is more likely to occur with increasing
sample size. However, the CIT is in general more fallable to reporting a
robust causal association for the wrong direction of causality. When
\(d > 0\) we find that in most cases the MR method has greater power to
obtain evidence for causality than CIT, and always obtains the correct
direction of causality. The CIT, unlike the Steiger test for MR, is able
to distinguish the ``non-causal model'' from the ``causal model''
(Methods, Figure 5b), but it is evident that measurement error will
often lead the CIT to identify the causal model as true, when in fact
the underlying model is this non-causal model.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation and gene expression levels between 458 putative
associations. We found that the causal direction commonly goes in both
directions (Figure 6a), but assuming no or equal measurement error, DNA
methylation levels were the predominant causal factor
(\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
was 3.92 (5\%-95\% quantiles 1.08 - 37.11). We then went on to predict
the causal directions of the associations for varying levels of
systematic measurement error for the different platforms. Figure 6a
shows that the conclusions about the direction of causality between DNA
methylation and gene expression are very sensitive to measurement error.

We performed two sample MR (28) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (41) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even in estimating associations
between `omic' variables, which are considered to be low level
phenotypes, it is important to use causal inference methods over
observational associations to infer causal effect sizes.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis are relatively well explored (44). Our analysis extends this to
related methods such as CIT that are widely used in predominantly 'omic
data. These methods are indeed susceptible to the same problem as
standard mediation based analysis, and specifically we show that as
measurement error in the (true) exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.
